---
title: "CABQ Transit Peer Review"
author: "Brendon Gray"
date: "May 2020 and revised June 2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(forcats)
library(scales)
library(stringr)
```

```{r, include=FALSE}
library(tidyverse) 
library(knitr)
library(viridis)

theme_minimal <- function(base_size = 11, base_family = "",
                          base_line_size = base_size / 22,
                          base_rect_size = base_size / 22) {
  # Starts with theme_bw and remove most parts
  theme_bw(
    base_size = base_size,
    base_family = base_family,
    base_line_size = base_line_size,
    base_rect_size = base_rect_size
  ) %+replace%
    theme(
      axis.ticks        = element_blank(),
      legend.background = element_blank(),
      legend.key        = element_blank(),
      panel.background  = element_blank(),
      panel.border      = element_rect(linetype = "solid", color = "grey", fill = NA),
      strip.background  = element_blank(),
      plot.background   = element_blank(),
      plot.caption = element_text(hjust = 0, vjust = 1),

      complete = TRUE
    )
}

theme_minimal_60 <- function(base_size = 11, base_family = "",
                          base_line_size = base_size / 22,
                          base_rect_size = base_size / 22) {
  # Starts with theme_bw and remove most parts
  theme_bw(
    base_size = base_size,
    base_family = base_family,
    base_line_size = base_line_size,
    base_rect_size = base_rect_size
  ) %+replace%
    theme(
      axis.text.x       = element_text(angle = 60),
      axis.ticks        = element_blank(),
      legend.background = element_blank(),
      legend.key        = element_blank(),
      panel.background  = element_blank(),
      panel.border      = element_blank(),
      panel.grid.major  = element_blank(),
      strip.background  = element_blank(),
      plot.background   = element_blank(),
      complete = TRUE
    )
}


theme_minimal_blank.x <- function(base_size = 11, base_family = "",
                          base_line_size = base_size / 22,
                          base_rect_size = base_size / 22) {
  # Starts with theme_bw and remove most parts
  theme_bw(
    base_size = base_size,
    base_family = base_family,
    base_line_size = base_line_size,
    base_rect_size = base_rect_size
  ) %+replace%
    theme(
      axis.title        = element_blank(),
      axis.ticks.x      = element_blank(),
      legend.background = element_blank(),
      legend.key        = element_blank(),
      plot.title        = element_text(hjust = 0),
      plot.subtitle     = element_text(hjust = 0),
      panel.background  = element_blank(),
      panel.border      = element_blank(),
      panel.grid.major  = element_blank(),
      panel.grid.minor  = element_blank(),
      axis.ticks        = element_blank(),
      strip.background  = element_blank(),
      plot.background   = element_blank(),
      axis.line.x.bottom = element_line(color = "red"),
      complete = TRUE
    )
}


theme_classic_transit <- function(base_size = 11, 
                                  # base_family = "sans",
                          base_line_size = base_size / 22,
                          base_rect_size = base_size / 22) {
  # Starts with theme_bw and remove most parts
  theme_classic(
    base_size = base_size,
    # base_family = base_family,
    base_line_size = base_line_size,
    base_rect_size = base_rect_size
  ) %+replace%
    theme(
      axis.title        = element_blank(),
      # axis.line.y.left  = element_blank(),
      legend.background = element_blank(),
      legend.key        = element_blank(),
      # plot.title        = element_text(hjust = 0),
      # plot.subtitle     = element_text(hjust = 0),
      panel.background  = element_blank(),
      panel.border      = element_blank(),
      panel.grid.major  = element_blank(),
      panel.grid.minor  = element_blank(),
      strip.background  = element_blank(),
      plot.background   = element_blank(),
      complete = TRUE
    )
}


```

```{R, include=FALSE}
# Load the raw data.

service_2019 <- read_excel("2019Service.xlsx")
service_2019_df <- service_2019 %>%
  select(
    "NTD ID",
    "Agency Name",
    "Time Period",
    "Mode",
    "TOS",
    "Vehicles/Passenger Cars Operated in Maximum Service",
    "Vehicles/Passenger Cars Available for Maximum Service",
    "Actual Vehicle/Passenger Car Miles",
    "Actual Vehicle/Passenger Car Revenue Miles",
    "Actual Vehicle/Passenger Car Deadhead Miles",
    "Actual Vehicle/Passenger Car Hours",
    "Actual Vehicle/Passenger Car Revenue Hours",
    "Actual Vehicle/Passenger Car Deadhead Hours",
    "Passenger Miles",
  ) %>%
  rename(
    ntd_id = "NTD ID",
    agency_name = "Agency Name",
    time_period = "Time Period",
    mode = "Mode",
    tos = "TOS",
    vechicles_op_max_service = "Vehicles/Passenger Cars Operated in Maximum Service",
    vehicles_avail_max_service = "Vehicles/Passenger Cars Available for Maximum Service",
    actual_vehicle_car_miles = "Actual Vehicle/Passenger Car Miles",
    actual_vehicle_car_revenue_miles = "Actual Vehicle/Passenger Car Revenue Miles",
    actual_vehicle_car_deadhead_miles = "Actual Vehicle/Passenger Car Deadhead Miles",
    actual_vehicle_car_hours = "Actual Vehicle/Passenger Car Hours",
    actual_vehicle_car_revenue_hours = "Actual Vehicle/Passenger Car Revenue Hours",
    actual_vehicle_car_deadhead_hours = "Actual Vehicle/Passenger Car Deadhead Hours",
    passenger_miles = "Passenger Miles",
  ) %>%
  filter(
    time_period == "Annual Total",
    !mode %in% c("CR", "DR", "DT", "FB", "VP"),
    tos == "DO",
    ntd_id %in% c(80005, 60019, 60006, 90033, 90032, 
                  80001, 90045, 80006, 90019, 50050, 
                  70005, 70035, 60017, 70002, 60018, 
                  90027, 90004)
    ) %>%
  as.data.frame(
  ) %>%
  mutate_at(., 
    "ntd_id", as.numeric
  )
service_2019_df

op_costs_2019 <- read_excel("2019OperatingExpenses.xlsx")

op_costs_2019_df <- op_costs_2019 %>%
  select(
    "NTD ID",
    "Agency Name",
    "Mode",
    "TOS",
    "Operating Expense Type",
    "Total Operating Expenses",
    "ADA Related Expenses"
  ) %>%
  rename(
    ntd_id = "NTD ID",
    agency_name = "Agency Name",
    mode = "Mode",
    tos = "TOS",
    operating_expense_type = "Operating Expense Type",
    total_operating_expenses = "Total Operating Expenses"
  ) %>%
  filter(
      operating_expense_type == "Total",
      !mode %in% c("CR", "DR", "DT", "FB", "VP"),
      tos == "DO",
      ntd_id %in% c(80005, 60019, 60006, 90033, 90032, 
                    80001, 90045, 80006, 90019, 50050, 
                    70005, 70035, 60017, 70002, 60018, 
                    90027, 90004)
  ) %>%
  as.data.frame(
  ) %>%
  mutate_at(., 
    "ntd_id", as.numeric
  )

ntd_metrics <- read.csv("NTD_Metrics.csv")
ntd_metrics_df <- ntd_metrics %>%
  select(
    Agency,
    City,
    NTD.ID,
    Agency.VOMS,
    Mode,
    TOS,
    Mode.VOMS,
    Fare.Revenues.per.Unlinked.Passenger.Trip,
    Fare.Revenues.per.Total.Operating.Expense..Recovery.Ratio.,
    Cost.per..Hour,
    Passengers.per.Hour,
    Cost.per.Passenger,
    Cost.per.Passenger.Mile,
    Fare.Revenues.Earned,
    Unlinked.Passenger.Trips,
    Vehicle.Revenue.Hours,
    Vehicle.Revenue.Miles
  ) %>%
  rename(
    agency_name = Agency,
    city = City,
    ntd_id = NTD.ID,
    agency_voms = Agency.VOMS,
    mode = Mode,
    tos = TOS,
    mode_voms = Mode.VOMS,
    fare_revenues_per_unlinked_passenger_trip = Fare.Revenues.per.Unlinked.Passenger.Trip,
    fare_revenues_per_total_operating_expense_recover_ratio = Fare.Revenues.per.Total.Operating.Expense..Recovery.Ratio.,
    cost_per_hour = Cost.per..Hour,
    passengers_per_hour = Passengers.per.Hour,
    cost_per_passenger = Cost.per.Passenger,
    cost_per_passenger_mile = Cost.per.Passenger.Mile,
    fare_revenues_earned = Fare.Revenues.Earned,
    unlinked_passenger_trips = Unlinked.Passenger.Trips,
    vehicle_revenue_hours = Vehicle.Revenue.Hours,
    vehicle_revenue_miles = Vehicle.Revenue.Miles    
  ) %>%
  filter(
      !mode %in% c("CR", "DR", "DT", "FB", "VP"),
      tos == "DO",
      ntd_id %in% c(80005, 60019, 60006, 90033, 90032, 
                    80001, 90045, 80006, 90019, 50050, 
                    70005, 70035, 60017, 70002, 60018, 
                    90027, 90004)
  ) %>%
  as.data.frame(
  ) %>%
  mutate_at(., 
    c("ntd_id"), as.numeric
  )

ntd_metrics_df

# join dfs

service_costs <- inner_join(service_2019_df, op_costs_2019_df)
service_costs_metrics <- inner_join(service_costs, ntd_metrics_df, by = c("ntd_id", "mode")) %>%
  rename(
    agency_name = agency_name.x,
    tos = tos.x
  ) %>%
  select(
    -agency_name.y,
    tos.y
  )

service_costs_metrics$agency_name <- as.factor(service_costs_metrics$agency_name)
service_costs_metrics

all_service_metrics_fin <- service_costs_metrics %>%
  mutate(
    agency_name = recode_factor(agency_name, 
                                "Indianapolis and Marion County Public Transportation" = "Indianapolis",
                                "City of El Paso" = "El Paso",
                                "Central Oklahoma Transportation and Parking Authority" = "Oklahoma City",
                                "Metropolitan Tulsa Transit Authority" = "Tulsa",
                                "City of Albuquerque" = "Albuquerque",
                                "Transit Authority of Omaha" = "Omaha",
                                "Kansas City Area Transportation Authority" = "Kansas City",
                                "Utah Transit Authority" = "Salt Lake City", 
                                "Denver Regional Transportation District" = "Denver",
                                "Golden Empire Transit District" = "Bakersfield", 
                                "Sacramento Regional Transit District" = "Sacramento",
                                "City of Fresno" = "Fresno")
  ) %>%
  mutate_all(funs(gsub(",","",.))
  ) %>%
  mutate_at(., c("fare_revenues_per_unlinked_passenger_trip", "cost_per_hour", "cost_per_passenger", "cost_per_passenger_mile", "fare_revenues_earned"), funs(str_remove_all(., "[$]"))
  ) %>%
  mutate_at(., 
    c("ntd_id", "agency_voms", "mode_voms", "fare_revenues_per_total_operating_expense_recover_ratio", 
      "passengers_per_hour", "unlinked_passenger_trips", "vehicle_revenue_hours", "vehicle_revenue_miles",
      "total_operating_expenses", "vechicles_op_max_service", "vehicles_avail_max_service", "actual_vehicle_car_miles",
      "actual_vehicle_car_revenue_miles", "actual_vehicle_car_deadhead_miles", "actual_vehicle_car_hours", "actual_vehicle_car_revenue_hours",
      "actual_vehicle_car_deadhead_hours", "passenger_miles", "fare_revenues_per_unlinked_passenger_trip", "cost_per_hour", "cost_per_passenger", "cost_per_passenger_mile", "fare_revenues_earned"), 
    as.numeric
  ) 


str(all_service_metrics_fin)

    
```

# Key Findings

- Albuquerque transit is more efficient than many peers.

- Fare revenues are much lower than other peer systems. 

A high-level peer review was conducted to compare Albuquerque transit’s performance relative to a group of peer systems. Performance was evaluated during 2018. 

A total of 12 comparison cites were selected. Population and geography were used as a guide in selecting this peer group. The peer cities are:

- Austin

- Bakersfield

- Colorado Springs

- El Paso

- Fresno

- Indianapolis

- Kansas City*

- Oklahoma City

- Omaha

- Sacramento*

- Tulsa

- Tucson

Some cities in the comparison group are much more populous than Albuquerque, such as Austin, Sacramento, Kansas City, and Indianapolis. Cities with an asterisk include multiple agencies within the region. A brief visualization comparing 12 larger systems is presented on page 6.

These statistics demonstrate Albuquerque meets or exceeds the average performance of its peers. Albuquerque tends to have a more efficient system than those in this peer group. The city has the third highest riders per capita, for example. Likewise, although Albuquerque’s operating costs and level service provision are about at the peer group average, the city’s efficiency with respect to cost and service provision is considerably higher than the peer group average.

This analysis is a simple representation that does not include many factors, such as cities’ demographics. No statistic perfectly captures a comparison between any two transit systems.

```{R, echo=FALSE, message=FALSE, warning=FALSE}
#, fig.width = 10
# annual revenue hours
rev_hours <- all_service_metrics_fin %>%
  group_by(
    agency_name
  ) %>%
  summarize(
    total_vehicle_rev_miles = sum(actual_vehicle_car_revenue_miles)/1000
  ) %>%
  mutate(
    agency_name = fct_reorder(agency_name, total_vehicle_rev_miles),
    avg_rev_miles = mean(total_vehicle_rev_miles)
  )
 
avg_rev_miles <- rev_hours$avg_rev_miles
  ggplot(data=rev_hours, aes(x = total_vehicle_rev_miles, y = agency_name, fill=factor(ifelse(agency_name=="Albuquerque","abq","nonabq")))) +
  geom_col(stat="identity", width = 0.70, position = position_dodge(), alpha = 1) +
  scale_fill_manual(name = "agency_name", values=c("#FFDD9C","#A8B7DF")) +
  geom_vline(aes(xintercept = avg_rev_miles, color = "Average"), linetype = "dashed") +
  labs(title = "Annual Revenue Hours",
       subtitle = "In '000s, 2019",
       y = "",
       caption = "Data from National Transit Database, 2019") +
  guides(fill=FALSE) +
  scale_x_continuous(labels = scales::comma) +
  theme_classic_transit() +
  theme(legend.title=element_blank())

```

# Demographics

The average service population of the peer group is about 706,000. Albuquerque is about 40,000 less populous than the average. Austin, TX has a population roughly twice as large as the average. Throughout the comparison, note that Austin is an outlier in this regard. Albuquerque is one of 5 cities in the southwest; the remaining are either on the west coast or in the Midwest. Comparisons across geography add another layer of complexity, as some systems have been operating for longer than Albuquerque’s. MSA population was not included in this analysis. 

# Trips
Albuquerque has a high number of unlinked passenger trips per capita, relative to the peer group. The per capita average is 9.7. Albuquerque’s 14.6 puts it well above that average, competing with cities like El Paso, TX and Fresno, CA.


# Service
Peers averaged 5.09 million revenue hours while Albuquerque provided just 4.04 million. Austin is a notable outlier within the peer group; removing that transit agency puts the peer group average at 4.36 million, just above Albuquerque. 
The amount of service provided and the number of annual trips represents the system productivity. Compared to the peer group, Albuquerque transit has high productivity. The city is about 60% more productive than the peer group average. Note that while the transit system in Austin offers the most number of revenue hours, it has one of the lowest productivity scores. 
Trips per revenue hour is not a perfect measure of productivity. A more accurate measure would take into consideration other demographic factors, like service population.


# Expenses
Operating costs were measured based off the cost of operating motor bus and demand response service only. These are the reference comparisons with Albuquerque.
The peer group averaged \$38.6 million in operating expenses in 2018. Albuquerque’s operating costs were roughly equal to the peer average, at \$42.8 million.  
As seen on the next page, Albuquerque’s \$106 cost per unit of service was above the peer group average of \$83. The city’s cost effectiveness, measured as the operating cost per trip, was \$4.43, well below the average of \$5.81. 
Albuquerque is, again, near the peer group average with overall operating costs and in the upper quarter with effectiveness measures.

```{R, echo=FALSE, message=FALSE, warning=FALSE}
# operating expenses
summ_op_expenses <- all_service_metrics_fin %>%
  group_by(
    agency_name
  ) %>%
  summarize(
    tot_op_expenses = sum(total_operating_expenses)/1000,
    op_expense_per_upt = tot_op_expenses/(sum(unlinked_passenger_trips)/1000),
    op_expense_per_rev_hour = tot_op_expenses/(sum(vehicle_revenue_hours)/1000)
  ) %>%
  mutate(
    avg_op_expenses = mean(tot_op_expenses),
    avg_op_expense_per_upt = mean(op_expense_per_upt),
    avg_op_expense_per_rev_hour = mean(op_expense_per_rev_hour)
  )

op_expenses <- summ_op_expenses %>%
  mutate(
    agency_name = fct_reorder(agency_name, tot_op_expenses, .desc = TRUE)
  )

op_expenses_per_upt <- summ_op_expenses %>%
  mutate(
    agency_name = fct_reorder(agency_name, op_expense_per_upt, .desc = TRUE)
  )

op_expenses_per_rev_hour <- summ_op_expenses %>%
  mutate(
    agency_name = fct_reorder(agency_name, op_expense_per_rev_hour, .desc = TRUE)
  )
 
avg_op_expenses <- summ_op_expenses$avg_op_expenses
avg_op_expense_per_upt <- summ_op_expenses$avg_op_expense_per_upt
avg_op_expense_per_rev_hour <- summ_op_expenses$avg_op_expense_per_rev_hour


ggplot(data=op_expenses, aes(x = tot_op_expenses, y = agency_name, fill=factor(ifelse(agency_name=="Albuquerque","abq","nonabq")))) +
  geom_col(stat="identity", position = position_dodge(), alpha = 1) +
  scale_fill_manual(name = "agency_name", values=c("#FFDD9C","#A8B7DF")) +
  geom_vline(aes(xintercept = avg_op_expenses, color = "Average"), linetype="dashed") +
  labs(title = "Annual Operating Costs",
       subtitle = "In '000s, 2019",
       y = "",
       caption = "Data from National Transit Database, 2019") +
  guides(fill=FALSE) +
  scale_x_continuous(labels = scales::comma) +
  theme_classic_transit() +
  theme(legend.title=element_blank())
  
  
ggplot(data=op_expenses_per_upt, aes(x = op_expense_per_upt, y = agency_name, fill=factor(ifelse(agency_name=="Albuquerque","abq","nonabq")))) +
  geom_col(stat="identity", position = position_dodge(), alpha = 1) +
  scale_fill_manual(name = "agency_name", values=c("#FFDD9C","#A8B7DF")) +
  geom_vline(aes(xintercept = avg_op_expense_per_upt, color = "Average"), linetype="dashed") +
  labs(title = "Operating Cost per Trip",
       subtitle = "2019",
       y = "",
       caption = "Data from National Transit Database, 2019") +
  guides(fill=FALSE) +
  scale_x_continuous(labels = scales::comma) +
  theme_classic_transit() +
  theme(legend.title=element_blank())


ggplot(data=op_expenses_per_rev_hour, aes(x = op_expense_per_rev_hour, y = agency_name, fill=factor(ifelse(agency_name=="Albuquerque","abq","nonabq")))) +
  geom_col(stat="identity", position = position_dodge(), alpha = 1) +
  scale_fill_manual(name = "agency_name", values=c("#FFDD9C","#A8B7DF")) +
  geom_vline(aes(xintercept = avg_op_expense_per_rev_hour, color = "Average"), linetype="dashed") +
  labs(title = "Operating Cost per Revenue Hour",
       subtitle = "2019",
       y = "",
       caption = "Data from National Transit Database, 2019") +  guides(fill=FALSE) +
  scale_x_continuous(labels = scales::comma) +
  theme_classic_transit() +
  theme(legend.title=element_blank())


# 
  
```

# Fares
Fare revenue was compared across the peer group. Although Albuquerque was at or above the peer group average for most of the above metrics, Albuquerque is behind with respect to fare revenue. The city’s \$3.2 m in fare revenue is well below the \$5.3 m peer group average. 
Comparing fare efficiency– i.e. fares per trip – still puts Albuquerque at the bottom of the sample. The city had the lowest fare revenues earned per trip, coming in 64% below the average. Albuquerque’s recovery ratio – the fare revenue earned per operating expense – was also the lowest of the peer group. Note cities with recovery ratios above 0.150 (Colorado Springs, Sacramento, Omaha, Bakersfield,  and Tucson) have relatively low service provision and service efficiency (page 2). This may indicate these systems operate under a different business model, which would explain the dramatic differences in recovery ratios.

```{R, echo=FALSE, message=FALSE, warning=FALSE}
# Revenues earned 

summ_revs_earned <- all_service_metrics_fin %>%
  filter(
    mode == "MB"
  ) %>%
  select(
    agency_name,
    fare_revenues_per_total_operating_expense_recover_ratio,
    fare_revenues_per_unlinked_passenger_trip
  ) %>%
  mutate(
    avg_fare_revenues_per_total_operating_expense_recover_ratio = mean(fare_revenues_per_total_operating_expense_recover_ratio),
    avg_fare_revenues_per_unlinked_passenger_trip = mean(fare_revenues_per_unlinked_passenger_trip),
  )

recov_ratio <- summ_revs_earned %>%
  mutate(
    agency_name = fct_reorder(agency_name, fare_revenues_per_total_operating_expense_recover_ratio)
  )

fare_revs_upt <- summ_revs_earned %>%
  mutate(
    agency_name = fct_reorder(agency_name, fare_revenues_per_unlinked_passenger_trip)
  )

avg_fare_revenues_per_total_operating_expense_recover_ratio <- summ_revs_earned$avg_fare_revenues_per_total_operating_expense_recover_ratio

avg_fare_revenues_per_unlinked_passenger_trip <- summ_revs_earned$avg_fare_revenues_per_unlinked_passenger_trip

ggplot(data=recov_ratio, aes(x = fare_revenues_per_total_operating_expense_recover_ratio, y = agency_name, fill=factor(ifelse(agency_name=="Albuquerque","abq","nonabq")))) +
  geom_col(stat="identity", position = position_dodge(), alpha = 1) +
  scale_fill_manual(name = "agency_name", values=c("#FFDD9C","#A8B7DF")) +
  geom_vline(aes(xintercept = avg_fare_revenues_per_total_operating_expense_recover_ratio, color = "Average"), linetype="dashed") +
  labs(title = "Fare Revenues Earned per Operating Expense - Recovery Ratio",
       subtitle = "2019",
       y = "",
       caption = "Data from National Transit Database, 2019") +
  guides(fill=FALSE) +
  scale_x_continuous(labels = scales::comma) +
  theme_classic_transit() +
  theme(legend.title=element_blank())


ggplot(data=fare_revs_upt, aes(x = fare_revenues_per_unlinked_passenger_trip, y = agency_name, fill=factor(ifelse(agency_name=="Albuquerque","abq","nonabq")))) +
  geom_col(stat="identity", position = position_dodge(), alpha = 1) +
  scale_fill_manual(name = "agency_name", values=c("#FFDD9C","#A8B7DF")) +
  geom_vline(aes(xintercept = avg_fare_revenues_per_unlinked_passenger_trip, color = "Average"), linetype="dashed") +
  labs(title = "Fare Revenues per UPT",
       subtitle = "2019",
       y = "",
       caption = "Data from National Transit Database, 2019") +
  guides(fill=FALSE) +
  scale_x_continuous(labels = scales::comma) +
  theme_classic_transit() +
  theme(legend.title=element_blank())

```

```{R, echo=FALSE, message=FALSE, warning=FALSE}
summ_revs_earned <- all_service_metrics_fin %>%
  filter(
    mode == "MB"
  ) %>%
  select(
    agency_name,
    fare_revenues_per_total_operating_expense_recover_ratio,
    fare_revenues_per_unlinked_passenger_trip
  ) %>%
  mutate(
    avg_fare_revenues_per_total_operating_expense_recover_ratio = mean(fare_revenues_per_total_operating_expense_recover_ratio),
    avg_fare_revenues_per_unlinked_passenger_trip = mean(fare_revenues_per_unlinked_passenger_trip),
  )

avg_fare_revenues_per_total_operating_expense_recover_ratio <- summ_revs_earned$avg_fare_revenues_per_total_operating_expense_recover_ratio
avg_fare_revenues_per_total_operating_expense_recover_ratio
avg_fare_revenues_per_unlinked_passenger_trip <- summ_revs_earned$avg_fare_revenues_per_unlinked_passenger_trip

recov_ratio <- summ_revs_earned %>%
  mutate(
    agency_name = fct_reorder(agency_name, fare_revenues_per_total_operating_expense_recover_ratio)
  )

fare_revs_upt <- summ_revs_earned %>%
  mutate(
    fare_per_upt_diffs = fare_revenues_per_total_operating_expense_recover_ratio - avg_fare_revenues_per_total_operating_expense_recover_ratio,
    is_diff_negative = ifelse(fare_per_upt_diffs < 0, "Below Average", "Above Average"),
    agency_name = fct_reorder(agency_name, fare_per_upt_diffs)
  ) 

fare_revs_upt

ggplot(data=fare_revs_upt, aes(x = fare_per_upt_diffs, y = agency_name, fill = is_diff_negative)) +
  geom_col(stat="identity", position = position_dodge(), alpha = 1) +
  scale_fill_manual(name="fare_per_upt_diffs", 
                    labels = c("Above Average", "Below Average"), 
                    values = c("#00ba38", "#f8766d")) +
  labs(title = "Fare Revenues Earned per Operating Expense - Recovery Ratio",
       subtitle = "2019",
       y = "",
       caption = "Data from National Transit Database, 2019") +
  theme_minimal_blank.x() +
  guides(fill=FALSE) +
  scale_x_continuous(labels = scales::comma)


ggplot(data=fare_revs_upt, aes(x = fare_per_upt_diffs, y = agency_name, label = agency_name)) +
  geom_point(stat='identity', fill = "black", size=10) +
  geom_segment(aes(y = fare_per_upt_diffs, 
                   x = 0, 
                   yend = agency_name, 
                   xend = fare_per_upt_diffs), 
                   color = "black") +
  geom_text(color = "white", size = 2) +
  labs(title = "Fare Revenues Earned per Operating Expense - Recovery Ratio",
       subtitle = "2019",
       y = "",
       caption = "Data from National Transit Database, 2019") +
  theme_minimal_blank.x() +
  guides(fill=FALSE) +
  scale_x_continuous(labels = scales::comma) +
  theme(legend.title=element_blank())






```

```{R, echo=FALSE, message=FALSE, warning=FALSE}
# operating cost per revenue hour
op_expenses <- all_service_metrics_fin %>%
  group_by(
    agency_name
  ) %>%
  summarize(
    tot_op_expenses = sum(total_operating_expenses)/1000
  ) %>%
  mutate(
    agency_name = fct_reorder(agency_name, tot_op_expenses, .desc = TRUE),
    avg_op_expenses = mean(tot_op_expenses)
  )
op_expenses 
avg_op_expenses <- op_expenses$avg_op_expenses

  ggplot(data=op_expenses, aes(x = tot_op_expenses, y = agency_name, fill=factor(ifelse(agency_name=="Albuquerque","abq","nonabq")))) +
  geom_col(stat="identity", position = position_dodge(), alpha = 1) +
  scale_fill_manual(name = "agency_name", values=c("#FFDD9C","#A8B7DF")) +
  geom_vline(xintercept = avg_op_expenses, linetype="dashed", color = "#DD934E") +
  labs(title = "Annual Operating Costs",
       subtitle = "In '000s, 2019",
       y = "",
       caption = "Data from National Transit Database, 2019") +
  theme_minimal_blank.x() +
  guides(fill=FALSE) +
  scale_x_continuous(labels = scales::comma)

```


# Large Systems Comparison
A sample of large systems were selected to compare against Albuquerque’s metrics. To simplify the comparison, three metrics were selected: service population, annual trips, and fare efficiency. Cities in orange have light rail service options (or the equivalent); cities in blue do not. As expected, cities with more population tend to have light rail service. Albuquerque has both a relatively small service population and a lower number of annual trips provided. Note cities who service similarly populous areas (Dallas, Portland, and Phoenix) all have below average fare efficiency. Albuquerque has comparable fare efficiency to this subset of systems. 







